---
title: "ggplot_intro"
author: "Matthew Kehoe"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## bring in data from previous file
```{r}
source(here::here("data/day3_objects.R"))
```

## Basic plotting: create scatter plot of life expectancy against gdpPercap
```{r}
ggplot(data = gapminder07) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp)) +
  labs(title = "Relationship between life expectancy and GDP per capita in 2007", x="GDP per capita", y="Life Expectancy")
```

## Exercise: create scatterplot of nat log gdpPercap as a funcation of nat log of pop
```{r}
ggplot(data = gapminder07) +
  geom_point(mapping = aes(x = log(pop), y = log(gdpPercap))) +
  labs(title = "log_gdpPercap vs. log_pop from the gapminder dataset in 2007", x="log_pop", y="log_gdpPercap")
```

## Preparing your data. Example: energy generated over time
```{r}
#plot column chart of total energy generated over time
# steps: choose long-format generation data, manipulate data frame to calculate total output per datatime, pipe manipulated dataframe into ggplot

long_gen %>%
  group_by(datetime) %>%
  summarize(output=sum(output)) %>%
  ggplot() +
    geom_col(aes(x=datetime, y=output)) +
    labs(title = "Total energy generated, by hour", x="Hour", y="Output (MW)")
```

## Exercise
```{r}
#plot column chart of hydroelectric power generated over time
generation %>%
  select(datetime,contains("hydro")) %>%
  melt(id.vars = "datetime", variable.name = "source", value.name = "usage") %>%
  group_by(datetime) %>%
  summarize(output=sum(usage)) %>%
  ggplot() +
    geom_col(aes(x=datetime, y=output), color = "light blue") +
    labs(title = "Total output from hydro sources, by hour", x="Hour", y="Output (MW)", caption = "Source: California Department of Energy") +
    scale_x_datetime(date_labels = "%H:%M", date_breaks = "12 hours")
```

## Multiple geoms in one plot
```{r}
# #plot generation of large hydro with line and a smooth line on top of the same relationship
# long_gen %>%
#   ggplot() +
#     geom_line(aes(x=datetime, y))
```

##Exercise
```{r}
#create column chart showing total output per source. change color of columns to "dark red".
#add horizontal line indication the mean output across all sources
long_gen %>%
  group_by(source) %>%
  summarize(total_output=sum(output)) %>%
  ggplot() +
    geom_col(aes(x=source, y=total_output), fill = "coral2") +
    geom_hline(aes(yintercept = mean(total_output)), color = "coral1") +
    theme(axis.text.x = element_text(angle=45, hjust=1, color = "coral1"))

```

## vizualizing grouped data
```{r}
#create a line plot of energy output over time, with separate lines for each source
long_merged_energy %>%
ggplot() +
  geom_line(aes(x=datetime, y=output, group=source, col=source))
  
```

## Example
```{r}
#create line plot comparing generation of wind, solar, and geothermal energy over time
long_gen %>%
  filter(source == "wind" | source == "solar" | source == "geothermal") %>%
  ggplot() +
    geom_line(aes(x=datetime, y=output, group=source, col=source ), size=1.15) +
    labs(title="Energy Generation over time, by source", x="Hour", y="Output (MW)", subtitle="Hourly data from September 3-9, 2018", caption="Source: California Department of Energy")
```

```{r}
#column chart of energy use by day grouped by source
long_merged_energy %>%
  mutate(date=lubridate::date(datetime)) %>%
  group_by(date, source) %>%
  summarize(output=sum(output)) %>%
  ggplot()+
    geom_col(aes(x=date, y=output, group=source, fill=source), position="dodge") +
    labs(title="Source of Energy Generation, by day", x="date", y="output (MW)")
```

## Exercise
```{r}
grouped <- merge(long_merged_energy, regroup, by.x = "source", by.y = "type")

grouped %>%
  group_by(group)%>%
  summarise(group_output=sum(output))%>%
  ggplot()+
    geom_line(aes(x=datetime, y=group_output, group=group, col=group))
```


## Exercise2
```{r}
#visualize average output for each hour of the day, grouped by group
grouped %>%
  mutate(hour=lubridate::hour(datetime)) %>%
  group_by(group, hour) %>%
  summarize(mean_group_output = mean(output)) %>%
  ggplot() +
    geom_line(aes(x=hour, y=mean_group_output, group=group, col=group)) +
    labs(title="Average Hourly Output by Source Group", x="Time of day", y="Mean Hourly Output", caption="Source: California Dept. of Energy", subtitle="Data Collected September 3rd-9th, 2018")

grouped %>%
  mutate(hour=lubridate::hour(datetime)) %>%
  group_by(group, hour) %>%
  summarize(mean_group_output = mean(output)) %>%
  ggplot() +
    geom_col(aes(x=hour, y=mean_group_output, group=group, fill=group), position="fill") +
    labs(title="Average Hourly Output by Source Group", x="Time of day", y="Mean Proportion of Total Hourly Output", caption="Source: California Dept. of Energy", subtitle="Data Collected September 3rd-9th, 2018")

grouped %>%
  mutate(hour=lubridate::hour(datetime)) %>%
  group_by(group, hour) %>%
  summarize(mean_group_output = mean(output)) %>%
  ggplot() +
    geom_line(aes(x=hour, y=mean_group_output, group=group, col=group)) +
    labs(title="Average Hourly Output by Source Group", x="Time of day", y="Mean Hourly Output", caption="Source: California Dept. of Energy", subtitle="Data Collected September 3rd-9th, 2018")+
    facet_wrap(~group, scales="free")
```

## Facet wrap with sources rather than groups
```{r}
grouped %>%
  ggplot() +
    geom_line(aes(x=datetime, y=output, group=source, col=group))+
    labs(title="Output by Source", x="Date", y="Output (MW)", caption="Source: California Dept. of Energy", subtitle="Data Collected September 3rd-9th, 2018")+
    facet_wrap(~source, scales="free")+
    theme(axis.text.x = element_text(angle=45, hjust=1), legend.position = "bottom") +
    scale_color_brewer(palette="Set2", name="Group of Energy source")
```

