---
title: "Intro to Loops, Conditionals, and Functions"
author: "Matthew Kehoe"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data", "gapminder5.csv")

gapminder <- read.csv(here::here("data", "gapminder5.csv"))
```

## Converting factors from dataset into characters
```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
```

## Calculate Means of life expectancy by country
```{r}
obs <- 1:nrow(gapminder)

for (i in obs){
  gapminder[i, "gdp"] <- gapminder[i, "pop"] * gapminder[i, "gdpPercap"]
  
}
```

## Try with from 1 to length, rather than with obs
```{r}
for (i in 1:length(gapminder$country)){
  gapminder[i, "gdp1"] <- gapminder[i, "pop"] * gapminder[i, "gdpPercap"]
  i = i+1
}
```


## Try this on your own:
```{r}
for (i in obs){
  gapminder[i, "log_gdpPercap"] <- log(gapminder[i, "gdpPercap"])
  gapminder[i, "log_pop"] <- log(gapminder[i, "pop"])
}
```

##Avoid loops if possible
```{r}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
all(gapminder$vec_log_gdpPercap == gapminder$log_gdpPercap)
```

## Means of life expectancy over years
```{r}
years <- unique(gapminder$year)

for (i in years){
  mean_le <- mean(gapminder$lifeExp[gapminder$year == i], na.rm = TRUE)
  mean_le <- round(mean_le, digits = 2)
  print(paste0(i, ": ", mean_le))
}
```

## Find mean value of life expectancy over continents
```{r}
cont <- unique(gapminder$continent)

for (i in cont){
  mean_cont <- mean(gapminder$lifeExp[gapminder$continent == i], na.rm = TRUE)
  mean_cont <- round(mean_cont, digits = 2)
  print(paste0(i, ": ", mean_cont))
}
```

## Mean by continent over years
```{r}
for (i in years){
  print(paste0("Year: ", i))
  for (j in cont){
    mean_cur <- mean(gapminder$lifeExp[gapminder$continent == j & gapminder$year == i], na.rm = TRUE)
    mean_cur <- round(mean_cur, digits = 2)
    print(paste0("   ", j, ": ", mean_cur))
  }
}
```

## Has the gap in life expectancy between continents  decreased over time?
```{r}
for (i in years){
  mean_vec <- NULL
  for (j in cont){
    mean_1 <- mean(gapminder$lifeExp[gapminder$continent == j & gapminder$year == i], na.rm = T)
    mean_vec <- cbind(mean_vec, mean_1)
  }
  stand <- round(sd(mean_vec), digits = 2)
  print(paste0(i,": ",stand))
}
```

## While Loops
## Find Sd of life expectancy for years before 1987
```{r}
i = years[1]
while (i < 1987){
  sd_lf <- sd(gapminder$lifeExp[gapminder$year == i], na.rm = T)
  print(paste0(i, ": ", sd_lf))
  i <- i+5
}
```

## Standard Deviation for life expectancy for each year between 1987 and 2002 (inclusive)
```{r}
y <- 1987
while (y <= 2002){
  sd_1 <- sd(gapminder$lifeExp[gapminder$year == y])
  print(paste0(y, ": ", round(sd_1, digits = 2)))
  y = y+5
}
```

## IF ELSE loops:
```{r}
#set.seed(10) #set a seed so that the "random number" is always the same. IE set seed to 1 for 1967
random_year <- sample(years, 1)

if (random_year > 1977){
  print(random_year)
}
```

## Use of if else:
```{r}
if(random_year < 1977){
  print(random_year)
} else{
  print("Sorry, the random year was later than 1977")
}
```

## Which continents have a mean life expectancy greater than 70?
```{r}
threshold = 70
for(i in unique(gapminder$continent)){
  tmp <- mean(gapminder$lifeExp[gapminder$continent == i])
  
  if (tmp < threshold){
    print(paste0(i," has a life expectancy LESS than 70 years"))
  } else{
    print(paste0(i, " has a life expectancy GREATER than 70 years"))
  }
}
```

## Write for loop that prints mean pop for years greater than or equal to 1987
```{r}
for (i in years){
  if (i < 1987){
    print("The year is prior to 1987. No mean calculated")
  } 
  else{
    print(paste0("The mean life expectancy for ", i, ": ", round(mean(gapminder$lifeExp[gapminder$year == i]), digits = 2)))
  }
}
```

## Writing functions: write function to print value of selected varaible in the gapminder dataset
```{r}
get_value <- function(y = "1987", count = "Hungary", var = "population"){
  print(paste0("The ", var, " of ", count, " in ", y, " was: ", gapminder[[var]][gapminder$country == count & gapminder$year == y]))
}

get_value()
gapminder$pop[gapminder$country == "Hungary" & gapminder$year == 1987]
```

## Function to report mean, median, min, max of life expectancy for a continent
```{r}
get_num <- function(cont = "Asia"){
  meanC <- mean(gapminder[[cont]])
  medianC <- median(gapminder$continent[cont])
  minC <- min(gapminder[[cont]])
  maxC <- max(gapminder[[cont]])
  print(c("Country:", cont,
          "\nMean Life Expectancy:", meanC,
          "\nMedian Life Expectancy:", medianC,
          "\nMinimum Life Expectancy:", minC,
          "\nMaximum Life Expectancy:", maxC))
}

get_num()
```

## answer
```{r}
report_stats <-
  function(df, variable, continent) {
    var <- gapminder[[variable]][gapminder$continent == continent]
    min_le <- min(var)
    max_le <- max(var)
    cat("Continent:", continent,
        "\nMinimum Life expectancy:", min_le,
        "\nMaximum Life expectancy:", max_le)
  }

report_stats(gapminder, "lifeExp", "Asia")
```
## Add this portion to practice pushing and pulling
